---
title: "ANTs longitudinal cortical thickness statistics: Region of interest version"
output: pdf_document
---

# Objective
Use regions of interest gained by multi-atlas labeling (MAL) to perform longitudinal statistics.

```{r fig.width=8, fig.height=4, echo=FALSE}
library(ANTsR)
library(plyr)
library(ggplot2)
library(date)
library(lme4)
library(nlme)
library(lmerTest)
library(packHV)
library(lattice)
library(visreg)
library(pheatmap)
verbose<-1
basedir<-"/Users/stnava/Downloads/LongitudinalX/"
setwd(basedir)
malf<-antsImageRead("pennTemplate/labels/antsMalfLabels.nii.gz",3)
data("DesikanKillianyTourville",package='ANTsR')
outdir<-paste(basedir,'FTD_ROI_Long',sep='')
dir.create(outdir, showWarnings = FALSE)
thicknessStudy<-FALSE
##############################
##############################
```

Define study demographic parameters.  
We use the demographics file to construct
the expected image file prefix.
```{r fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE,error=FALSE}
### Most important thing - load a csv
### location of subject directories
demog<-read.csv(paste(basedir,'./FTD_Long_Nov_2014.csv',sep='/'))
expectedImagePrefix<-rep("NA",nrow(demog))
haveSubjectTimePoint<-rep(FALSE,nrow(demog))
for ( i in 1:nrow(demog) )
  {
  myid<-demog$INDDID[i]
  rawdate<-as.character(demog$MRIDate[i])
  rawdatesplit<-unlist(strsplit(rawdate,'/'))
  locyear<-rawdatesplit[3]
  locmon<-rawdatesplit[1]
  locday<-rawdatesplit[2]
  expectedImagePrefix[i]<-paste(myid,'_',locyear,locmon,locday,sep='')
  findfn<-Sys.glob(paste(basedir,'/Murray/',myid,
    '/',expectedImagePrefix[i],"*",sep=''))
  if ( length(findfn) > 0) haveSubjectTimePoint[i]<-TRUE
  }
if ( verbose > 1 )  print(sum(haveSubjectTimePoint))
demog<-cbind(demog,expectedImagePrefix=expectedImagePrefix,
  haveSubjectTimePoint=haveSubjectTimePoint)
```

Now find out how many subjects we have.
```{r countdata, fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE}
type1<-"*OASIS30_gm.nii.gz"
type2<-"*CorticalThickness.nii.gz"
# count the subjects 
subjectct<-0
haveSubject<-rep(FALSE,length(demog$expectedImagePrefix))
ct<-1
for ( x in demog$expectedImagePrefix  )
  {
  temp<-Sys.glob(paste(basedir,"Murray/*/*/*/",x,type1,sep=''))
  temp2<-Sys.glob(paste(basedir,"Murray/*/*/*",x,type2,sep=''))
  if ( length(temp) > 0 & length(temp2) > 0 & 
         demog$DaysApart[ct] < (365*2.1) )
    {
    subjectct<-subjectct+1
    haveSubject[ct]<-TRUE
    }
  if ( length(temp) > 1 ) stop(paste("multi-match error",x))
  ct<-ct+1
  }
```

Now compute the csv summary files from the ROI images.  
We compute volume, thickness and surface area for each 
region that is available in the DKT labels that we 
usually use.  The labels come from 
[Arno's Mindboggle](http://mindboggle.info/data.html).
```{r getcsvs, fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE}
nrois<-nrow(DesikanKillianyTourville)
if ( ! exists("myroidf") )
  {
  myroidf<-matrix(rep(rep(0,nrois),subjectct),nrow=subjectct)
  myroidf<-data.frame(myroidf)
  myroidfThickness<-myroidf
  myroidfArea<-myroidf
  colnames(myroidf)<-DesikanKillianyTourville$LabelName
  ct<-1
  for ( x in demog$expectedImagePrefix[haveSubject]  )
    {
    if ( verbose ) print(x)
    temp<-Sys.glob(paste(basedir,"Murray/*/*/*/",x,type1,sep=''))
    temp2<-Sys.glob(paste(basedir,"Murray/*/*/*",x,type2,sep=''))
    roi<-antsImageRead(temp,3)
    thk<-antsImageRead(temp2,3)
    # index the images by the colnames of the data frame
    # not fast but is clear and safe
    spc<-antsGetSpacing(roi)
    volelt<-spc[1]*spc[2]*spc[3]
    tempfn<-paste(outdir,"/",x,'.csv',sep='')
    if ( ! file.exists(tempfn) )
      ImageMath(3,tempfn,"LabelStats",roi,thk)
    roivals<-read.csv(tempfn)
    colct<-1
    for ( anat in colnames(myroidf) )
      {
      num<-as.numeric(DesikanKillianyTourville$LabelNum[
        DesikanKillianyTourville$LabelName==anat])
      if ( sum(roivals$label == num ) == 1 )
        {
        #FIXME - someone else should check this!!!
        ctval<-roivals$count[ roivals$label == num  ]
        myvol<-ctval*volelt
        mythk<-roivals$mass[ roivals$label == num  ]
        myarea<-myvol/(mythk/ctval)
        myroidf[ct,colct ]<-myvol
        myroidfThickness[ct, colct]<-mythk
        myroidfArea[ct, colct]<-myarea
        }
      colct<-colct+1
      }
    ct<-ct+1
    }
  } 
```

Let's quickly investigate these results.
```{r vizcorrs}
  myroidf[ myroidf == 0 ]<-NA
  myroidfThickness[ myroidfThickness == 0 ]<-NA
  myroidfArea[ myroidfArea == 0 ]<-NA
  meanthk<-rowMeans(myroidfThickness,na.rm=T)
  meanvol<-rowMeans(myroidf,na.rm=T)
  meanare<-rowMeans(myroidfArea,na.rm=T)
  print(cor.test(meanthk,meanvol))
#  pheatmap(antsrimpute(cor(data.matrix(myroidfArea))))
```

Run through the results with mixed effects modeling.

```{r mixeff, fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE}
# make an image to save results
makevizimg<-function( m1 ) {
  vizimg<-antsImageClone(m1);
  vizimg[ vizimg > 0 ]<-0
  return(vizimg)
  }
subdemog<-demog[haveSubject,]
mydx<-as.factor( subdemog$ClinicalPhenotype1 )
mydx = factor(mydx,levels(mydx)[c(2,1,3,4,5)])
subdemog<-cbind(subdemog,DX=mydx)
pvsBV<-rep(NA,ncol(myroidf))
vizBV<-makevizimg( malf )
pvsLV<-rep(NA,ncol(myroidf))
vizLV<-makevizimg( malf )
pvsN<-rep(NA,ncol(myroidf))
vizN<-makevizimg( malf )
pvsSV<-rep(NA,ncol(myroidf))
vizSV<-makevizimg( malf )
for ( jj in 1:ncol(myroidf) )
  {
  imging<-myroidf[,jj]
  if ( thicknessStudy ) 
    imging<-myroidfThickness[,jj]
  if ( ! any(is.na(imging)) )
  {
  locdf<-data.frame( img=imging, Sex=subdemog$Sex,
    Age=subdemog$AgeatMRI, 
    Edu=as.factor(subdemog$EducationCat),
    Prof=as.factor(subdemog$ProfessionalCat_1), 
    dTime=as.numeric(subdemog$DaysApart),
    DX=mydx,
#    TDP=as.factor(subdemog$LikelyTDP), 
    sId=as.factor(subdemog$INDDID) )
  myform<-as.formula( img ~ 1 + Age + dTime:DX +
    Sex + Edu + Prof+ (1|sId) )
  mdl<-lmer( myform , data=(locdf))
  if ( verbose == 1 ) print(summary(mdl))
  smdl<-summary(mdl)$coeff
  pvsBV[jj]<-smdl[7,5]
  pvsLV[jj]<-smdl[8,5]
  pvsN[jj]<-smdl[9,5]
  pvsSV[jj]<-smdl[10,5]
  labnum<-as.numeric( DesikanKillianyTourville$LabelNumber[jj] )
  vizBV[ malf == labnum ]<-1.0-pvsBV[jj]
  vizLV[ malf == labnum ]<-1.0-pvsLV[jj]
  vizN[ malf == labnum ]<-1.0-pvsN[jj]
  vizSV[ malf == labnum ]<-1.0-pvsSV[jj]
  locallabel<-colnames(myroidf)[jj]
  if ( verbose > 1 )  print(paste(locallabel))
  p<-ggplot(locdf, aes( x = dTime ,y = img, group = sId)) +  geom_line() + geom_smooth(aes(group = 1), method="lm", size = 2, se = T) + geom_point(aes(colour = factor(DX)))

#  p<-ggplot(data=locdf , aes( x = dTime,
#    y = img, group = sId )) +
#    geom_line()  + geom_point() + ggtitle(locallabel)
  plot(p)
#  Sys.sleep(1)
  }
}

```


Report some results
```{r reportRegions}
regionStats<-function( pvs, nm ) {
  print(nm)
  print(paste(colnames(myroidf)[which.min(pvs)],
    p.adjust(pvs)[which.min(pvs)]))
  qv<-p.adjust(pvs,'BH')
  cth<-max( pvs[ qv <= 0.05 ] , na.rm=T)
  print(paste("CriticalThresh",cth))
  return(cth)
}
bvth<-regionStats( pvsBV, "BV")
lvth<-regionStats( pvsLV, "LV")
nth<-regionStats( pvsN, "N")
svth<-regionStats( pvsSV, "SV")
```

Set up visualization: see `antsSurfViz.sh`.
```{r vizsurf}
# write the images then do some system calls to antsSurf
antsImageWrite(vizBV,'/tmp/malf_vizBV.nii.gz')
antsImageWrite(vizLV,'/tmp/malf_vizLV.nii.gz')
antsImageWrite(vizN,'/tmp/malf_vizN.nii.gz')
antsImageWrite(vizSV,'/tmp/malf_vizSV.nii.gz')
```

