---
title: "ANTs longitudinal cortical thickness statistics: Region of interest version"
output: pdf_document
---

# Objective
Use regions of interest gained by multi-atlas labeling (MAL) to perform longitudinal statistics.

```{r fig.width=8, fig.height=4, echo=FALSE}
library(ANTsR)
library(plyr)
library(ggplot2)
library(date)
library(lme4)
library(nlme)
library(lmerTest)
library(packHV)
library(lattice)
library(visreg)
verbose<-1
##############################
##############################
```

Define study demographic parameters.  
We use the demographics file to construct
the expected image file prefix.
```{r fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE,error=FALSE}
### Most important thing - load a csv
### location of subject directories
basedir<-"/Users/stnava/Downloads/LongitudinalX/antslongct/"
demog<-read.csv(paste(basedir,'../FTD_Long_Nov_2014.csv',sep='/'))
expectedImagePrefix<-rep("NA",nrow(demog))
haveSubjectTimePoint<-rep(FALSE,nrow(demog))
for ( i in 1:nrow(demog) )
  {
  myid<-demog$INDDID[i]
  rawdate<-as.character(demog$MRIDate[i])
  rawdatesplit<-unlist(strsplit(rawdate,'/'))
  locyear<-rawdatesplit[3]
  locmon<-rawdatesplit[1]
  locday<-rawdatesplit[2]
  expectedImagePrefix[i]<-paste(myid,'_',locyear,locmon,locday,sep='')
  findfn<-Sys.glob(paste(basedir,'/',myid,
    '/',expectedImagePrefix[i],"*",sep=''))
  if ( length(findfn) > 0) haveSubjectTimePoint[i]<-TRUE
  }
if ( verbose > 1 )  print(sum(haveSubjectTimePoint))
demog<-cbind(demog,expectedImagePrefix=expectedImagePrefix,
  haveSubjectTimePoint=haveSubjectTimePoint)
```

Define study imaging parameters
```{r fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE}
studytype<-c('jacobian','thickness')[1]
outprefix<-paste('/Users/stnava/Downloads/LongitudinalX/temp/FTD_Stats_',studytype,sep='')
# get all of these paths / filenames right ....
maskfn<-"/Users/stnava/Downloads/LongitudinalX/pennTemplate/masks/templateBrainMaskProbability.nii.gz"
testmaskfn<-"/Users/stnava/Downloads/LongitudinalX/pennTemplate/labels/antsMalfLabels.nii.gz"
if ( ! file.exists(maskfn) ) if ( verbose > 1 )  print(paste(maskfn,'doesnt exist'))
if ( ! file.exists(testmaskfn) ) if ( verbose > 1 )  print(paste(testmaskfn,'doesnt exist'))
temimg<-antsImageRead(maskfn,3)
smoothingparam<-2.0 # smoothing before stats
```

Now do the preprocessing

Get the csv summary files.
```{r fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE}
csvs<-c()
csvs<-rep(NA,nrow(demog))
ncolslist<-rep(NA,nrow(demog))
ct<-1
for ( x in demog$expectedImagePrefix  )
  {
  temp<-Sys.glob(paste(basedir,"*/*/",x,"*MPR*brainvols.csv",sep=''))
  if ( length(temp) > 0 )
    {
    csvs[ct]<-temp
    tempcsv<-read.csv(temp)
    ncolslist[ct]<-ncol(tempcsv)
    }
  if ( length(temp) > 1 ) stop(paste("multi-match error",x))
  ct<-ct+1
  }
if ( sd( ncolslist[ !is.na(ncolslist) ] ) > 0 )
  stop("different n columns in some files")
csvdf<-data.frame(matrix( rep(NA, ncolslist[ !is.na(ncolslist) ][1]*nrow(demog) ),
  nrow=nrow(demog) ))
colnames(csvdf)<-colnames(tempcsv)
ct<-1
for ( x in demog$expectedImagePrefix  )
  {
  temp<-Sys.glob(paste(basedir,"*/*/",x,"*MPR*brainvols.csv",sep=''))
  if ( length(temp) > 0 )
    {
    csvs[ct]<-temp
    tempcsv<-read.csv(temp)
    csvdf[ct,]<-tempcsv[1,]
    ncolslist[ct]<-ncol(tempcsv)
    }
  ct<-ct+1
  }
myimagingdf <-data.frame( demog, csvdf  )
```

```{r derka, fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE}
mydelt<-rep(NA,nrow(demog))
ct<-2 # start here b/c we check one back
while ( ct <= nrow(myimagingdf ) )
  {
  if ( demog$INDDID[ct] == demog$INDDID[ct-1] &
       ! is.na(demog$INDDID[ct]) )
    {
    basect<-which( demog$INDDID == demog$INDDID[ct] )[1]
    allct<-which( demog$INDDID == demog$INDDID[ct] )
    demog$AgeatMRI[ allct ]<-demog$AgeatMRI[basect]
    mydelt[basect]<-0
    dt1<-as.character( demog$MRIDate[ct] )
    dt2<-as.character( demog$MRIDate[basect] )
    mydelt[ct]<-as.date(dt1,order='mdy')-
                as.date(dt2,order='mdy')
    }
  ct<-ct+1
  }
```

```{r derka2, fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE}
myimagingdf<-cbind(myimagingdf,INDDID=demog$INDDID,timeSinceBaseline=mydelt)
```


```{r fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE}
p<-ggplot(data=myimagingdf , aes( x = timeSinceBaseline, y = BVOL,
  group = INDDID )) +   geom_line()  + geom_point()
plot(p)
```

```{r firstplot, fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE}
for ( jj in 19:22 )
  {
  imging<-myimagingdf[,jj]
  myform<-as.formula( paste("imging~timeSinceBaseline+(1|INDDID)" ) )
  mxf<-lmer(myform,data=myimagingdf)
  smxf<-summary(mxf)
  if ( verbose == 1 )  print(smxf)
  locallabel<-colnames(myimagingdf)[jj]
  if ( verbose > 1 )  print(paste(locallabel))
  p<-ggplot(data=myimagingdf , aes( x = timeSinceBaseline,
    y = imging, group = INDDID )) +
    geom_line()  + geom_point() + ggtitle(locallabel)
  plot(p)
}
````


```{r derka4, fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE}
bigct<-1
countup<-0
haveimg<-rep(0,nrow(demog))
for ( loopid in demog$expectedImagePrefix )
  {
  locid<-as.character( loopid )
  if (  myimagingdf$haveSubjectTimePoint[bigct] > 0 &
      ! is.na( loopid ) )
    {
    myid<-as.character(myimagingdf$INDDID[bigct])
    if ( studytype =='thickness')
      { # FIXME - should be more robust
      ipttrn<-glob2rx(paste("*",locid,"*Thickness.nii.gz",sep=''))
      warpfn<-Sys.glob(paste(basedir,"*/*/*",locid,
        "*SubjectToGroupTemplateWarp.nii.gz",sep=''))
      }
    if ( studytype =='jacobian')
      { # FIXME - should be more robust
      ipttrn<-glob2rx(paste("*",locid,"*ToTemplateLogJacobian.nii.gz",sep=''))
  afffn<-Sys.glob(paste(basedir,myid,'/',myid,"*SingleSubjectTemplate/T_templateSubjectToTemplate0GenericAffine.mat",sep='')) # FIXME - should be like imagelistfn
  warpfn<-Sys.glob(paste(basedir,myid,'/',"*SingleSubjectTemplate/T_templateSubjectToTemplate1Warp.nii.gz",sep=''))# FIXME - should be like imagelistfn
      }
      imagelistfn<- paste(basedir,list.files(path=basedir,
        pattern = ipttrn ,recursive=T),sep='/')
      if ( file.exists(warpfn[1]) & file.exists(imagelistfn[1]) )
      {
      ct<-1
      imagelist<-list()
      for ( jfn in imagelistfn ) {
        if ( studytype == 'jacobian' ) txlist<-c( warpfn, afffn )
        if ( studytype != 'jacobian' ) txlist<-c( warpfn[ct] )
        wout<-paste(outprefix,locid,'warped.nii.gz',sep='' )
        if ( verbose > 1 )  print(wout)
        countup<-countup+1
        haveimg[bigct]<-1
        if ( ! file.exists(wout) )
          {
          jimg<-antsImageRead( jfn, 3)
          wimg<-antsApplyTransforms( fixed=temimg,
            moving=jimg, transformlist=txlist)
          SmoothImage(3,wimg,smoothingparam,wimg)
          antsImageWrite( wimg , wout )
          } else {
          if ( verbose > 1 )  print(paste(locid,wout,'done'))
    #      wimg<-antsImageRead( wout, 3 )
          }
        ct<-ct+1
      }
    }
    }
  bigct<-bigct+1
  }
#
#if ( exists("brain") )
#      plotANTsImage( brain, functional=list(corrs,ncorrs),
#        threshold="0.8x1",slices="40x190x5",color=c("red","blue"),
#        outname=ofn, alpha=0.25)
#
```

Run the voxel-wise stats.

```{r stats,fig.width=8, fig.height=4, echo=TRUE,warnings=FALSE,message=FALSE}
wimgfns<-Sys.glob(paste(outprefix,"*","warped.nii.gz",sep=''))
lab<-0
tmask<-antsImageRead(testmaskfn,3)
bmask<-antsImageClone( tmask )
bmask[ bmask > 0 ]<-1
bmask[ bmask <= 0 ]<-0
if ( lab > 0 ) {
roimask<-antsImageClone( tmask )
roimask[ roimask != as.numeric(lab) ]<-0
roimask[ roimask == as.numeric(lab) ]<-1
} else roimask<-antsImageClone(bmask)
wout<-paste(outprefix,'bmask.nii.gz',sep='' )
antsImageWrite(bmask,wout)
wout<-paste(outprefix,'roimask.nii.gz',sep='' )
antsImageWrite(roimask,wout)
sum(roimask==1)
imglist<-imageFileNames2ImageList( wimgfns, 3 )
imging<-imageListToMatrix( imglist, roimask )
subdf<-demog[haveimg==1,]
fid<-as.factor( subdf$INDDID )
tsb<-myimagingdf$timeSinceBaseline[haveimg==1]
quadmodel<-stats::poly( tsb, 1 )

doMixedEffects<-FALSE
if (  doMixedEffects )
{
# eanat on baseline images
eanat<-sparseDecom(
  inmatrix=imging[subdf$timeSinceBaseline==0,],
  inmask=roimask, nvecs=5,
  its=3, sparseness=0.1, smooth=0.0, cthresh=150)
pmat<-abs(imageListToMatrix(eanat$eig,roimask))
vecsums<-apply(pmat,FUN=sum,MARGIN=1)
pmat<-pmat/vecsums
projs<-( imging %*%  t(pmat) )
mydx<-as.factor( subdf$ClinicalPhenotype1 )
mydx = factor(mydx,levels(mydx)[c(2,1,3,4,5)])
for ( j in 1:ncol(projs) ) {
  locdf<-data.frame( img=projs[,j], Sex=subdf$Sex,
    Age=subdf$AgeatMRI, Edu=as.factor(subdf$EducationCat),
    Prof=as.factor(subdf$ProfessionalCat_1), dTime=tsb,
    DX=mydx,
    TDP=as.factor(subdf$LikelyTDP), sId=as.factor(fid) )
  myform<-as.formula( img ~ 1 + Age + dTime:TDP +
    Sex + Edu + Prof+ (1|sId) )
  mdl<-lmer( myform , data=(locdf))
  if ( verbose == 1 ) print(summary(mdl))
  getStat<-function( mdlIn ) {
    summdl<-summary(mdlIn)
    smk<-4
    if ( dim(summdl$coeff)[2] > (smk) ) {
#      return( (summdl$coeff[smk,4])*(-1) )
     return( (summdl$coeff[smk,4]-summdl$coeff[smk+1,4])*(1) )
     } else return(0)
    }
  basestat<-getStat( mdl )
  pct<-0
  for ( i in 1:100 ) {
    sampdf<-locdf
    for ( x in unique(locdf$sId) ) {
      loctdp<-locdf$TDP[ locdf$sId ==x ]  
      if ( !is.na(loctdp[1]) )
        sampdf$TDP[ locdf$sId == x ]<-sample(levels(locdf$TDP))[1]
      }
    mdl<-lmer( myform , data=sampdf)
    permstat<-getStat(mdl)
    if ( permstat > basestat ) pct<-pct+1
    print(paste(i,pct,permstat))
  }
}
} else { # do two level stats
# first get individual level stats for each subject
  mysubs<-unique(subdf$INDDID[tsb==0])
  matvals<-rep(0,ncol(imging)*length(mysubs))
  firstlevelmat<-matrix( matvals, nrow=length(mysubs) )
  rownames(firstlevelmat)<-rep("",length(mysubs))
  ct<-1
  for ( x in mysubs ) {
    locmat<-imging[subdf$INDDID==x,  ]
    loctime<-stats::poly(tsb[subdf$INDDID==x],1)
    mdl<-lm( locmat  ~ loctime )  
    mdl<-bigLMStats( mdl, 0.00001 )
    firstlevelmat[ct,]<-((mdl$beta))
    rownames(firstlevelmat)[ct]<-x
    ct<-ct+1
    }
  secondleveldemog<-subdf[tsb==0, ]
  mydx<-as.factor( secondleveldemog$ClinicalPhenotype1 )
  mydx = factor(mydx,levels(mydx)[c(2,1,3,4,5)])
  secondleveldemog<-cbind(secondleveldemog,DX=mydx)
  myform<-as.formula( firstlevelmat ~ 1 + AgeatMRI +
      LikelyTDP + Sex + ProfessionalCat_1 )
  fullmdl<-lm( myform , data = secondleveldemog )
  bigfullmdl<-bigLMStats( fullmdl, 1.e-5 )
  betaout<-antsImageClone( bmask )
  pv<-bigfullmdl$beta.pval["LikelyTDP",]
  qv<-p.adjust(pv,method='BH')
  qv[ is.na(qv)]<-1
  betaout[roimask==1]<-1.0-qv
  wout<-paste(outprefix,'qv.nii.gz',sep='' )
  antsImageWrite(betaout,wout)
  vizind<-which.min(qv)
  locdf<-data.frame( img=firstlevelmat[,vizind], Sex=secondleveldemog$Sex,
    Age=secondleveldemog$AgeatMRI,
    Edu=as.factor(secondleveldemog$EducationCat),
    Prof=as.factor(secondleveldemog$ProfessionalCat_1),
    DX=secondleveldemog$DX,
    TDP=as.factor(secondleveldemog$LikelyTDP) )
  myform<-as.formula( img ~ 1 + Age +
      TDP + Sex + Prof  )
  fullmdl<-lm( myform , data = locdf )
  visreg(fullmdl,'Age')
  visreg(fullmdl,'TDP')
  visreg(fullmdl,'Prof')
  doEanat<-FALSE
  if ( doEanat ) {
    eanat<-sparseDecom( inmatrix=firstlevelmat,
        inmask=roimask, nvecs=25, its=2,
        sparseness=0.005, smooth=0.0, cthresh=150)
    pmat<-abs(imageListToMatrix(eanat$eig,roimask))
    vecsums<-apply(pmat,FUN=sum,MARGIN=1)
    pmat<-pmat/vecsums
    projs<-( firstlevelmat %*%  t(pmat) )
    myform<-as.formula( img ~ 1 + AgeatMRI +
      LikelyTDP +
      Sex + ProfessionalCat_1 )
    for ( j in 1:ncol(projs))
      {
      img<-projs[,j]
      mdl<-lm( myform , data=(secondleveldemog))
      if ( verbose == 1 ) print(summary(mdl))
      }
    eseg<-eigSeg( roimask , eanat$eigenanatomyimages  )
    wout<-paste(outprefix,'eseg.nii.gz',sep='' )
    antsImageWrite(eseg,wout)
    }
}
```

Set up visualization: see `antsSurfViz.sh`.
